<center> <h1>The tutorial to use ChipSeq</h1> </center>

+ Run ChipSeq to perform one-setp analysis for ChipSeq related data 

```{r eval=FALSE}
dir.name="/media/H_driver/Aimin_project/"

file.sample="decoder.bySample.Mut_WT_2.rtf"
file.gff="Homo_sapiens.GRCh38.84.processed.sorted.4.JunctionSeq.flat.gff"

file.count="/QC.spliceJunctionAndExonCounts.forJunctionSeq.txt"

Re.example<-GetResultsFromJunctionSeq(dir.name,file.sample,file.count,file.gff)
 
```

+ Convert the results of differential usage analysis into gene based resutls

```{r eval=FALSE}
re.example.gene.based<-makeGeneWiseTable(Re.example,gene.list=unique(as.character(fData(Re.example)$geneID)))
```

+ Apply logistic regression model to identify bias factor
```{r eval=TRUE}
library(PathwaySplice)
data(mds)
re<-LRtestBias(mds)
```

+ Perform pathwaysplice in one step
```{r eval=TRUE}
library(PathwaySplice)
data(mds)
Example.Go.adjusted.by.exon<-Run_pathwaysplice(mds,ad="exon_SJ",sub_feature="E",0.05,genomeID="hg19",geneID="ensGene",gene_model=hg19.gene.model,method="Wallenius")
Example.Go.adjusted.by.exon<-Run_pathwaysplice(mds,ad="exon_SJ",sub_feature="E",0.05,genomeID="hg19",geneID="ensGene",gene_model=hg19.gene.model,method="Sampling")
Example.Go.unadjusted<-Run_pathwaysplice(mds,ad="exon_SJ",sub_feature="E",0.05,genomeID="hg19",geneID="ensGene",gene_model=hg19.gene.model,method="Hypergeometric")

```

+ If you are interested in other gene sets such as Canonical Pathways(CP),Transcription Factor Targets(TFT) and hallmark gene sets from http://software.broadinstitute.org/gsea/msigdb/collections.jsp, download these .gmt files, then perform the following steps:
```{r eval=TRUE}
gene.2.cat.hallmark.hg<-Gmt2GeneCat("/media/H_driver/Annotation/hg38/h.all.v5.1.symbols-1.gmt","/media/H_driver/Annotation/hg38/genes_table_02092016.csv")
```

+ Build up network based on the overlap between gene sets and visualize this network

```{r eval=TRUE}
data(mds)
Example.Go.adjusted.by.exon<-Run_pathwaysplice(mds,ad="exon_SJ",sub_feature="E",0.05,genomeID="hg19",geneID="ensGene",gene_model=hg19.gene.model,method="Wallenius")
re.w.adjusted<-enrichmentMap(Example.Go.adjusted.by.exon,n=5,SimilarityThreshold=0)
```
